---
title: "Generalizing princeBART results to external populations"
author: "princeBART authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalizing princeBART results to external populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)
# Helper: find a file in inst/extdata regardless of working directory
.extdata <- function(fname) {
  f <- system.file("extdata", fname, package = "princeBART")
  if (nzchar(f) && file.exists(f)) return(f)
  vdir <- tryCatch(
    dirname(normalizePath(knitr::current_input())),
    error = function(e) getwd()
  )
  f <- normalizePath(file.path(vdir, "..", "inst", "extdata", fname), mustWork = FALSE)
  if (file.exists(f)) return(f)
  stop("Cannot find extdata/", fname)
}
```


## Generalizing to External Populations

A key feature of princeBART is the ability to **generalize treatment effects** 
to an external population using `general_BART()`. This is useful when:

- You have a source study with detailed covariates and an instrument
- You want to estimate the Population Average Treatment Effect (PATE) in a 
  larger target population (e.g., a nationally representative survey)
- The target population may be missing some covariates collected in the source

The key identifying assumptions are:

1. **Conditional transportability**: The conditional treatment effect given X is 
   the same in the source study and target population. That is, once we condition 
   on observed covariates X, there are no remaining unmeasured effect modifiers 
   that differ between the two in a way that changes the effect.

2. **Included support**: Covariate profiles observed in the target population must 
   be adequately represented in the source data.

### Simulating an External Population

Let's simulate a large external "survey" population that:
1. Has different covariate distributions than the source study
2. Is missing the `income` variable (a common scenario with survey data)
3. Has complex survey features (PSUs and weights)

```{r simulate-external}
library(princeBART)
set.seed(123)
n_external <- 500

# External population overlaps with source but shifts slightly
external_data <- data.frame(
  age = rnorm(n_external, 43, 11),        # Slightly older on average
  education = rnorm(n_external, 11.5, 3.5) # Slightly lower education
  # Note: income is NOT available in external data
)
# Survey design: 100 PSUs with varying sizes
external_data$psu <- sample(1:100, n_external, replace = TRUE)
external_data$weight <- runif(n_external, 0.5, 2)  # Survey weights

# Define a target subpopulation (e.g., adults under 60)
external_data$eligible <- external_data$age < 60
```



### Generalizing to the External Population

Now we use `general_BART()` to:
1. Auto-detect and multiply impute missing variables (e.g., `income`) using BART
2. Compute instrument propensity e = P(Z|X) as a feature expansion
3. Predict potential outcomes Y(0) and Y(1) for external units
4. Estimate the PATE using Bayesian bootstrap for complex surveys


```{r general-bart, eval = FALSE}
fit <- readRDS(system.file("extdata", "fit_intro.rds", package = "princeBART"))

# Generalize treatment effects to external population
# Missing variables are auto-detected and imputed
pate_result <- general_BART(
  princebart_fit = fit,
  newdata = external_data,
  subpop = external_data$eligible,       # Target subpopulation
  psu = external_data$psu,               # PSU for survey inference
  weights = external_data$weight,        # Survey weights
  n_cores = 4,
  verbose = TRUE
)

# View results
print(pate_result)

# saveRDS(pate_result, file = "inst/extdata/pate_result.rds")
```

The output shows:
- **PATE**: The population average treatment effect estimate
- **95% CI**: Credible interval incorporating both posterior and survey uncertainty
- **N (subpop)**: Number of units in the target subpopulation

The `general_pate` object stores all necessary information (predicted outcomes, 
survey design, model components) for downstream sensitivity analyses.

### Understanding the Output

```{r pate-summary}
pate_result <- readRDS(.extdata("pate_result.rds"))

# Detailed summary
summary(pate_result)

# Access posterior draws for custom analyses
hist(pate_result$draws,
     main = "Posterior Distribution of PATE",
     xlab = "Treatment Effect")
abline(v = pate_result$pate, col = "red", lwd = 2)
abline(v = pate_result$ci, col = "red", lty = 2)
```

### Assessing the Support Assumption: Overlap Diagnostics

One key threat to valid generalization is **lack of included support**—if target 
covariate profiles are poorly represented in the source data, results may depend 
on extrapolation into under-represented regions of covariate space. To assess this, 
use `general_BART_overlap()` to compute the **generalizability overlap score**:

$$s = P(\text{complier}|X, \text{in source}) \times P(\text{in source}|X)$$

This measures how similar external units are to compliers in the source study and 
whether their covariate profiles are well-represented in the source.

For binary uptake fits, this is exactly the complier-based overlap diagnostic.
For ordinal uptake fits, princeBART uses an affected-unit analogue
\(W(0)-W(1)=1\); this ordinal overlap diagnostic is currently experimental.

```{r overlap-analysis, eval = FALSE}
# Compute overlap scores from the fitted general_pate object
overlap <- general_BART_overlap(
  object = pate_result,
  verbose = TRUE
)
```
```{r overlap-analysis-plot}
overlap <- readRDS(.extdata("overlap_result.rds"))

# View overlap metrics
head(overlap$overlap)

# Plot overlap diagnostics
plot_overlap(overlap)
```

The overlap plot shows the distribution of selection scores for:
- Source study compliers (blue)
- Target population units (red)

Good overlap means the distributions substantially overlap.

For ordinal uptake fits, this diagnostic uses an affected-unit analogue
\(W(0)-W(1)=1\) in place of binary compliers and should be treated as
experimental.

#### Conservative Support Sensitivity Check

If some target-population units have covariate profiles poorly represented in 
the source, you can assess sensitivity by recomputing the PATE under a conservative 
assumption for those units. This tests whether conclusions depend heavily on 
extrapolation:

```{r overlap-trimming, eval = FALSE}
# Identify and conservatively adjust for poorly supported target units
# Set treatment effect to zero for units with |e_s_tilde| > 2 (standardized score)
overlap_trimmed <- general_BART_overlap(
  object = pate_result,
  threshold = 2,           # Standardized score threshold
  overlap_value = "zero",  # Conservative: assume tau = 0 for poorly supported units
  verbose = TRUE
)
```

```{r overlap-trimming-results}
overlap_trimmed <- readRDS(.extdata("overlap_trimmed_result.rds"))

# Compare conservative PATE vs original
cat("Original PATE (all units):", round(pate_result$pate, 4), "\n")
cat("Conservative PATE (unsupported → 0):", round(overlap_trimmed$pate_trimmed, 4), "\n")
cat("Units with weak support:", overlap_trimmed$n_trimmed, "\n")
```

### Sensitivity Analysis for Transportability

Even with good overlap/support, the source and target populations may differ in 
**unmeasured effect modifiers**—covariates not captured in X. To assess sensitivity 
to this violation of conditional transportability, use `general_BART_transportability()` 
for a bounded weight-shift sensitivity analysis. This computes bounds on the PATE 
under the assumption that an unobserved effect modifier differs between populations 
by a specified amount.

Run the sensitivity analysis across a range of gamma values to produce a 
sensitivity curve. This shows how robust the PATE is to unmeasured effect 
modification across source-target differences of varying severity:

```{r sensitivity-curve, eval = FALSE}
# Sensitivity curve for multiple gamma values
gammas <- c(1.1, 1.25, 1.5, 2, 3)
sens_results <- lapply(gammas, function(g) {
  general_BART_transportability(pate_result, gamma = g)
})
```

```{r sensitivity-curve-results}
sens_results <- readRDS(.extdata("sensitivity_curve_results.rds"))

gammas <- c(1.1, 1.25, 1.5, 2, 3)
# Extract bounds
lower_bounds <- sapply(sens_results, function(x) x$lower$estimate)
upper_bounds <- sapply(sens_results, function(x) x$upper$estimate)

# Plot sensitivity curve
plot(gammas, upper_bounds, type = "l", col = "red", lwd = 2,
     ylim = range(c(lower_bounds, upper_bounds)),
     xlab = "Gamma (max weight ratio)", ylab = "PATE Bounds", 
     main = "Sensitivity to Unmeasured Effect Modifiers")
lines(gammas, lower_bounds, col = "blue", lwd = 2)
abline(h = 0, lty = 2, col = "gray")
abline(h = pate_result$pate, lty = 3, col = "dark gray", lwd = 1.5)
legend("topright", c("Upper bound", "Lower bound", "Null effect", "Point estimate"), 
       col = c("red", "blue", "gray", "dark gray"), lty = c(1, 1, 2, 3), lwd = c(2, 2, 1, 1.5))
```

### Summary: Two Robustness Checks for Generalization

The vignette has illustrated two distinct robustness checks that address the two 
key identifying assumptions:

**Support Sensitivity Check** (via overlap diagnostics and conservative adjustment)
- **Assumption tested**: Included support
- **Question**: Are we extrapolating into covariate regions poorly represented 
  in the source?
- **Evidence**: Threshold selected based on standardized selection score; PATE 
  recomputed assuming zero effect for poorly supported target units
- **Interpretation**: Large drop in PATE suggests strong dependence on 
  extrapolation; modest drop suggests robustness to extrapolation risk

**Transportability Sensitivity Check** (via weight-shift bounds)
- **Assumption tested**: Conditional transportability
- **Question**: How large would source-target differences in unmeasured effect 
  modifiers need to be to materially change the PATE?
- **Evidence**: Sensitivity curve showing PATE bounds across increasing values 
  of gamma (maximum weight ratio)
- **Interpretation**: If bounds remain tight across large gamma values and/or 
  exclude zero, results are robust to unmeasured effect modification; if bounds 
  widen dramatically or cross zero at small gamma, findings are fragile

Both checks are valuable for a complete assessment of generalizability.
