---
title: "Introduction to princeBART"
author: "princeBART authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to princeBART}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)
```

## Overview
 
princeBART implements **Principal Stratification with Bayesian Additive Regression Trees (BART)** for causal inference with endogenous treatments. It estimates treatment effects for compliers—units whose treatment status is affected by an instrument.

While the canonical application is noncompliance in randomized experiments, the framework applies broadly to any setting with:

- An **instrument Z** that affects treatment uptake but has no direct effect on outcomes
- An **endogenous treatment W** whose causal effect on Y is of interest
- **Covariates X** that may confound the Z-W or W-Y relationships



## The Setup

We observe:

- **Z**: Instrument (e.g., randomized assignment, geographic variation, policy change)
- **W**: Endogenous treatment/exposure of interest
- **Y**: Outcome
- **X**: Covariates

### Key Assumptions

1. **Conditional randomization**: $Z \perp\!\!\!\perp (Y(z,w), W(z)) | X$ — the instrument is as-good-as-random given covariates
2. **Exclusion restriction**: Z affects Y only through W
3. **Monotonicity**: No defiers (units for whom Z decreases W)

### Principal Strata

The population consists of three principal strata:

| Stratum | W if Z=0 | W if Z=1 | 
|---------|----------|----------|
| Compliers | 0 | 1 |
| Never-takers | 0 | 0 |
| Always-takers | 1 | 1 |

The key estimand is the **Complier Average Causal Effect (CACE)**, also known as the Local Average Treatment Effect (LATE)—the causal effect of W on Y for units whose treatment is affected by the instrument. princeBART also targets conditional complier effects given covariates, denoted CATE_C(x).

## Simulated Example Data

Let's create a simulated dataset with clear treatment effect heterogeneity
to demonstrate \code{segment_heterogeneity()}.

```{r simulate-data}
library(princeBART)
set.seed(42)
n <- 800

# Covariates
X <- data.frame(
  age = rnorm(n, 40, 10),
  education = rnorm(n, 12, 3),
  income = rnorm(n, 50000, 15000)
)

# Principal strata (latent)
# Probability of being a complier depends on covariates
p_complier <- plogis(-1.2 + 0.03 * X$age + 0.12 * X$education - 0.00001 * X$income)
p_always   <- plogis(-2 + 0.01 * X$age)
p_never    <- pmax(1 - p_complier - p_always, 0)
denom <- p_complier + p_never + p_always
p_complier <- p_complier / denom
p_never <- p_never / denom
p_always <- p_always / denom


strata <- sapply(1:n, function(i) {
  sample(c("complier", "never", "always"), 1,
         prob = c(p_complier[i], p_never[i], p_always[i]))
})

# Instrument (randomized, possibly conditional on X)
Z <- rbinom(n, 1, 0.5)

# Treatment uptake depends on strata and instrument
W <- Z
W[strata == "always"] <- 1L
W[strata == "never"]  <- 0L

# Heterogeneous complier treatment effect
tau <- 0.2 +
  0.15 * (X$education > 12) -
  0.10 * (X$age > 50) +
  0.08 * (X$income > 60000)

# Potential outcomes (only Y(W) is observed)
lin0 <- -0.6 + 0.01 * X$age + 0.04 * X$education + 0.00001 * X$income
Y0 <- plogis(lin0)
Y1 <- plogis(lin0 + tau)

Y <- rbinom(n, 1, ifelse(W == 1, Y1, Y0))

# Combine into data frame
simdata <- data.frame(X, Z = Z, W = W, Y = Y)

# True ATE among compliers
true_ate_c <- mean(tau[strata == "complier"])
```

## Basic Usage

### Formula Interface

princeBART uses a 2SLS-style formula: `Y ~ covariates | Z | W`

```{r formula-usage, eval = FALSE}

library(future)

# Set up parallel backend (works on all platforms)
plan(multisession, workers = 4)

# Fit the model
fit <- prince_BART(
  Y ~ age + education + income | Z | W,
  data = simdata,
  n_chains = 2,
  n_warmup = 100,
  n_samples = 200,
  instrument_overlap = c(0.1, 0.9),  # Crump et al. (2009) recommendation
  keep_trees = TRUE  # Save trees for generalization
)

```



## Extracting Treatment Effects

princeBART provides average effects for compliers (LATE-like effects) via summary() and coef(). 
These are mixed estimands—averages of conditional complier effects over the empirical covariate distribution—
and can optionally be restricted to treated compliers (treated_only = TRUE).

```{r estimands}
fit <- readRDS(system.file("extdata", "fit_intro.rds", package = "princeBART"))

# Print summary
print(fit)

# Summary shows strata distribution and treatment effect
summary(fit)

# Mixed ATE for Compliers (default)
coef(fit)
```

## Effect Heterogeneity by Segments

We can automatically find segments (subgroups) with different conditional complier
effects using a shallow regression tree on posterior mean CATE_C(x).
The segment-level summaries in res$effects correspond to subgroup-specific
average effects among compliers, obtained by averaging conditional effects
within each covariate-defined segment. Differences across segments highlight
departures from effect homogeneity and provide an interpretable summary of
treatment effect heterogeneity.
```{r heterogeneity, fig.width=8, fig.height=5, out.width="100%"}
res <- segment_heterogeneity(
  fit,
  min_compliers_bucket = 50,
  plot = TRUE,
  contrast = TRUE
)

res$contrast$summary
res$plot$diff

```

The segment-level summaries in res$effects correspond to subgroup-specific
average effects among compliers, obtained by averaging conditional effects
within each covariate-defined segment. Differences across segments highlight
departures from effect homogeneity and provide an interpretable summary of
treatment effect heterogeneity. Segments are constructed post hoc to summarize 
heterogeneity in the posterior distribution of CATE_C(x).
```{r heterogeneity_effects, fig.width=8, fig.height=5, out.width="100%"}
res$effects
res$plot$effect
```