---
title: "Prince BART with Ordinal Uptake"
author: "princeBART authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prince BART with Ordinal Uptake}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)
library(princeBART)
.extdata <- function(fname) {
  f <- system.file("extdata", fname, package = "princeBART")
  if (nzchar(f) && file.exists(f)) return(f)
  vdir <- tryCatch(
    dirname(normalizePath(knitr::current_input())),
    error = function(e) getwd()
  )
  f <- normalizePath(file.path(vdir, "..", "inst", "extdata", fname), mustWork = FALSE)
  if (file.exists(f)) return(f)
  stop("Cannot find extdata/", fname)
}
```

## Overview

While earlier vignettes focus on binary treatment uptake, many empirical
applications involve treatments that take multiple ordered values. This
vignette illustrates how `princeBART` can be used when the treatment is
**discrete with more than two levels**, using fertility (number of children) as
a motivating example.

We consider a setting in which:

* **Z**: current infecundity indicator (instrument)
* **W**: number of children ever born (treatment)
* **Y**: employment (outcome)
* **X**: baseline covariates such as age, education, and urban residence

The number of children is a **discrete, non-negative variable with a finite
range** (e.g., 0–8 children in the simulated data below). This differs from the
binary uptake case but fits naturally within the Prince BART framework.

For binary uptake, the IV estimand can be interpreted as the average effect of
switching from untreated to treated among compliers. With multivalued uptake,
the instrument can shift treatment by different magnitudes across units.
In `princeBART`, the primary ordinal contrast focuses on units with
`W(0) - W(1) = 1`, that is, units whose treatment is shifted by exactly one
unit by the instrument. In this fertility-style example, the corresponding
summary can be interpreted as the average impact of one additional child on
employment for women in that one-unit-shift principal stratum.

This example is inspired by applications where current infecundity is used as
an instrument for fertility and employment is the outcome of interest. See the
methodological and empirical application in:

https://arxiv.org/abs/2508.10787

Conceptually, this vignette combines two ideas illustrated separately in
earlier vignettes:

1. **Detecting treatment effect heterogeneity** using conditional effects among
  units with a one-unit treatment shift (`W(0) - W(1) = 1`).
2. **Generalizing those conditional effects** to a broader population by
   averaging over a target covariate distribution.

In this example, the target population is defined **within the same survey
dataset**, rather than through a separate external dataset.

## Setup

We observe:

* **Z**: binary instrument
* **W**: ordinal treatment taking values `0, 1, ..., J`
* **Y**: binary outcome
* **X**: baseline covariates

Our goal is to estimate conditional effects for one-unit treatment changes
(`W(0) - W(1) = 1`) and summarize how these effects vary across values of `X`.

## Identification Assumptions

The analysis relies on standard instrumental-variable assumptions adapted to the
multivalued-treatment setting.

1. **Conditional unconfoundedness of the instrument given X**
   After adjusting for observed covariates `X`, the instrument `Z` is assumed
   independent of the relevant potential outcomes.

2. **Exclusion restriction**
   The instrument affects the outcome only through the treatment `W`.

3. **Monotonicity**
  In this application, infecundity can only weakly reduce fertility, so
  `W(1) <= W(0)` under the coding where `Z = 1` indicates infecundity.

4. **Additional assumption for multivalued treatments**
  With multivalued treatments, identification additionally requires an
  assumption about the residual association between `W(0)` and `W(1)` after
  conditioning on observed covariates `X`.


## Simulated Example Data

We simulate a dataset inspired by fertility applications where current
infecundity affects the number of children ever born, and an additional child
may reduce employment in a heterogeneous way.

```{r simulate-data}
simulate_ps_ordinal <- function(
  n = 1000,
  max_w = 8,
  seed = 123
) {
  set.seed(seed)

  age <- pmin(pmax(round(rnorm(n, mean = 29, sd = 8)), 18), 45)

  education <- sample(
    c("none", "primary", "secondary+"),
    size = n,
    replace = TRUE,
    prob = c(0.25, 0.30, 0.45)
  )

  urban <- rbinom(n, 1, 0.45)

  sexually_active <- rbinom(
    n, 1,
    plogis(-3 + 0.18 * (age - 18) + 0.5 * urban)
  )

  educ_num <- c("none" = 0, "primary" = 1, "secondary+" = 2)[education]

  pz <- plogis(-6.2 + 0.14 * age + 0.4 * sexually_active - 0.15 * educ_num)
  Z <- rbinom(n, 1, pz)

  latent_w0 <- -1.5 +
    0.18 * (age - 18) +
    0.25 * educ_num -
    0.15 * urban +
    rnorm(n, sd = 1.0)

  W0 <- pmax(0, pmin(max_w, round(latent_w0)))

  delta <- rbinom(
    n, size = 2,
    prob = plogis(-1.3 + 0.15 * W0 + 0.02 * (age - 25))
  )

  W1 <- pmax(0, W0 - delta)

  W <- ifelse(Z == 1, W1, W0)

  tau_child <- ifelse(
    education == "none", -0.2,
    ifelse(education == "primary", -0.1, 0.00)
  )

  eta0 <- -0.7 +
    0.10 * educ_num +
    0.10 * urban +
    0.02 * (age - 25) -
    0.0008 * (age - 25)^2

  eta_y <- eta0 + tau_child * W
  py <- plogis(eta_y)

  Y <- rbinom(n, 1, py)

  data.frame(
    Y = Y,
    Z = Z,
    W = W,
    age = age,
    education = ordered(education,
      levels = c("none", "primary", "secondary+")
    ),
    urban = urban,
    W0_true = W0,
    W1_true = W1,
    pz_true = pz,
    tau_child_true = tau_child
  )
}

set.seed(42)
simdata <- simulate_ps_ordinal(n = 1000, seed = 42)

table(simdata$W)
mean(simdata$Z)
mean(simdata$Y)
```

## Basic Usage

### Formula Interface

`princeBART` uses a 2SLS-style formula:

`Y ~ covariates | Z | W`

```{r formula-usage, eval = FALSE}
library(princeBART)
library(future)
plan(multisession, workers = 4)

fit <- prince_BART(
  Y ~ education + age + urban | Z | W,
  rho = 0,
  data = simdata,
  n_chains = 2,
  n_warmup = 100,
  n_samples = 200,
  instrument_overlap = c(0.1, 0.9),
  keep_trees = TRUE
)

# saveRDS(fit, file = "inst/extdata/fit_ordinal.rds")
```

## Extracting Treatment Effects

`princeBART` summarizes effects among units with a one-unit instrument-induced
treatment shift (`W(0) - W(1) = 1`). These summaries average conditional
effects over the empirical covariate distribution.

```{r estimands}
fit <- readRDS(.extdata("fit_ordinal.rds"))

print(fit)

summary(fit)

coef(fit)
```
## Sensitivity to the multivalued-treatment assumption

With multivalued treatments, identification requires an assumption about the
association between the two potential treatment levels `W(0)` and `W(1)`.
This association is not identified from observed data because only one of the
two potential treatment values is observed for each individual.

In `princeBART`, this assumption is governed by `rho`, which controls residual
dependence in the latent bivariate uptake model.

* `rho = 0`: conditional independence between latent `W(0)` and `W(1)` given `X`
* `rho > 0`: units with higher baseline uptake tend to have higher uptake under `Z = 1`
* `rho < 0`: units with higher baseline uptake tend to have lower uptake under `Z = 1`

Because `rho` is not empirically identified, varying `rho` is a natural
sensitivity analysis for ordinal-uptake conclusions.

## Effect Heterogeneity by Segments

We can summarize heterogeneity by fitting a shallow regression tree to the
posterior mean conditional effects.

```{r heterogeneity}

res <- segment_heterogeneity(
  fit,
  min_compliers_bucket = 2,
  plot = TRUE,
  contrast = TRUE
)

res$contrast$summary
res$plot$diff
```

The segment summaries correspond to subgroup-specific averages of conditional
effects. Differences across segments highlight interpretable heterogeneity in
treatment effects.

```{r heterogeneity_effects}
res$effects
res$plot$effect
```
## Generalizing to a Broader Population

In earlier vignettes, generalization relied on a separate external dataset
containing the covariate distribution of the target population. Here we
illustrate a related setting where the target population is defined within the
same dataset.

Generalizing treatment effects requires that the conditional effects estimated
in the instrument-identified subgroup are valid for the target population once
we average over the target covariate distribution. This relies on two
conditions:
(i) **conditional transportability**, meaning that after adjusting for the
observed covariates (X), the conditional treatment effects do not differ
systematically between the estimation sample and the target population; and
(ii) **adequate overlap** in the covariate distributions so that the values of
(X) observed in the target population are represented in the estimation
sample.

Diagnostics and practical guidance for assessing these conditions are discussed
in the *Generalization with princeBART* vignette.

In this example, the target population is defined using a subpopulation indicator
within the same dataset. The generalization step then averages the conditional
effects estimated by `princeBART` over the covariate distribution of that
target group.

```{r simulate-external}
n <- nrow(simdata)

# Survey design: 100 PSUs with varying sizes
simdata$psu <- sample(1:100, n, replace = TRUE)

# Survey weights
simdata$weight <- runif(n, 0.5, 2)

# Define a target subpopulation (e.g., adults under 60)
simdata$eligible <- simdata$age < 60
```

```{r general-bart}

# Generalize treatment effects to the target population
pate_result <- general_BART(
  princebart_fit = fit,
  newdata = simdata,
  subpop = simdata$eligible,       # Target subpopulation
  psu = simdata$psu,               # PSU for survey inference
  weights = simdata$weight,        # Survey weights
  n_cores = 4,
  verbose = TRUE
)

# View results
print(pate_result)
```
