---
title: "Generalizing princeBART results to external populations"
author: "princeBART authors"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Generalizing princeBART results to external populations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  fig.width = 8,
  fig.height = 5,
  out.width = "100%"
)
```


## Generalizing to External Populations

A key feature of princeBART is the ability to **generalize treatment effects** 
to an external population using `general_BART()`. This is useful when:

- You have a source study with detailed covariates and an instrument
- You want to estimate the Population Average Treatment Effect (PATE) in a 
  larger target population (e.g., a nationally representative survey)
- The target population may be missing some covariates collected in the source

The key identifying assumption is that **X captures all effect heterogeneity**, 
meaning the conditional treatment effect given X is the same for compliers in 
the source study as in the general population.

### Simulating an External Population

Let's simulate a large external "survey" population that:
1. Has different covariate distributions than the source study
2. Is missing the `income` variable (a common scenario with survey data)
3. Has complex survey features (PSUs and weights)

```{r simulate-external}
library(princeBART)
set.seed(123)
n_external <- 500

# External population overlaps with source but shifts slightly
external_data <- data.frame(
  age = rnorm(n_external, 43, 11),        # Slightly older on average
  education = rnorm(n_external, 11.5, 3.5) # Slightly lower education
  # Note: income is NOT available in external data
)
# Survey design: 100 PSUs with varying sizes
external_data$psu <- sample(1:100, n_external, replace = TRUE)
external_data$weight <- runif(n_external, 0.5, 2)  # Survey weights

# Define a target subpopulation (e.g., adults under 60)
external_data$eligible <- external_data$age < 60
```



### Generalizing to the External Population

Now we use `general_BART()` to:
1. Auto-detect and multiply impute missing variables (e.g., `income`) using BART
2. Compute instrument propensity e = P(Z|X) as a feature expansion
3. Predict potential outcomes Y(0) and Y(1) for external units
4. Estimate the PATE using Bayesian bootstrap for complex surveys


```{r general-bart, eval = FALSE}
fit <- readRDS(system.file("extdata", "fit_intro.rds", package = "princeBART"))

# Generalize treatment effects to external population
# Missing variables are auto-detected and imputed
pate_result <- general_BART(
  princebart_fit = fit,
  newdata = external_data,
  subpop = external_data$eligible,       # Target subpopulation
  psu = external_data$psu,               # PSU for survey inference
  weights = external_data$weight,        # Survey weights
  n_cores = 4,
  verbose = TRUE
)

# View results
print(pate_result)
```

The output shows:
- **PATE**: The population average treatment effect estimate
- **95% CI**: Credible interval incorporating both posterior and survey uncertainty
- **N (subpop)**: Number of units in the target subpopulation

The `general_pate` object stores all necessary information (predicted outcomes, 
survey design, model components) for downstream sensitivity analyses.

### Understanding the Output

```{r pate-summary}
pate_result <- readRDS(system.file("extdata", "pate_result.rds", package = "princeBART"))


# Detailed summary
summary(pate_result)

# Access posterior draws for custom analyses
hist(pate_result$draws,
     main = "Posterior Distribution of PATE",
     xlab = "Treatment Effect")
abline(v = pate_result$pate, col = "red", lwd = 2)
abline(v = pate_result$ci, col = "red", lty = 2)
```

### Assessing Generalizability Overlap

A natural concern is whether the source study population is representative of 
the target. Use `general_BART_overlap()` to compute the **generalizability overlap score**:

$$s = P(\text{complier}|X, \text{in source}) \times P(\text{in source}|X)$$

This measures how similar external units are to compliers in the source study.

```{r overlap-analysis, eval = FALSE}
# Compute overlap scores from the fitted general_pate object
overlap <- general_BART_overlap(
  object = pate_result,
  verbose = TRUE
)

```
```{r overlap-analysis-plot}
overlap <- readRDS(system.file("extdata", "overlap_result.rds", package = "princeBART"))

# View overlap metrics
head(overlap$overlap)

# Plot overlap diagnostics
plot_overlap(overlap)
```

The overlap plot shows the distribution of selection scores for:
- Source study compliers (blue)
- Target population units (red)

Good overlap means the distributions substantially overlap.

#### Trimming for Overlap

If some external units have poor overlap with source compliers, you can trim 
them from the PATE estimate using a threshold on the standardized selection score:

```{r overlap-trimming, eval = FALSE}
# Trim observations with |e_s_tilde| > 2 (i.e., outliers)
overlap_trimmed <- general_BART_overlap(
  object = pate_result,
  threshold = 2,           # Standardized score threshold
  overlap_value = "zero",  # Set tau = 0 for trimmed units
  verbose = TRUE
)
```

```{r overlap-trimming-results}
overlap_trimmed <- readRDS(system.file("extdata", "overlap_trimmed_result.rds", package = "princeBART"))
# Compare trimmed vs original PATE
cat("Original PATE:", round(pate_result$pate, 4), "\n")
cat("Trimmed PATE: ", round(overlap_trimmed$pate_trimmed, 4), "\n")
cat("Units trimmed:", overlap_trimmed$n_trimmed, "\n")
```

### Sensitivity Analysis for Confounding

If you're concerned about unmeasured confounding, use `general_BART_transportability()` 
for weight-shift sensitivity analysis:

```{r sensitivity, eval = FALSE}
# Sensitivity analysis with gamma = 2
# This computes bounds on PATE under worst-case weight perturbations
sens <- general_BART_transportability(
  object = pate_result,
  gamma = 2,        # Allow weights to vary by factor of 2
  n_sample = 100,   # Number of posterior draws to use
  verbose = TRUE
)

```
```{r sensitivity_results}
sens <- readRDS(system.file("extdata", "sensitivity_result.rds", package = "princeBART"))
# View sensitivity bounds
cat("Gamma =", sens$gamma, "\n")
cat("Lower bound:", round(sens$lower$estimate, 4), 
    "[", round(sens$lower$ci[1], 4), ",", round(sens$lower$ci[2], 4), "]\n")
cat("Upper bound:", round(sens$upper$estimate, 4),
    "[", round(sens$upper$ci[1], 4), ",", round(sens$upper$ci[2], 4), "]\n")
```

The sensitivity analysis shows bounds on the PATE under worst-case weight 
perturbations. The `gamma` parameter controls the maximum ratio by which 
observation weights can be shifted—larger values allow for more severe 
confounding. If the bounds exclude zero, the effect is robust to confounding
of that magnitude.

#### Multiple Sensitivity Parameters

You can run sensitivity analysis for multiple gamma values to create a 
sensitivity curve:

```{r sensitivity-curve, eval = FALSE}
# Sensitivity curve for multiple gamma values
gammas <- c(1.1, 1.25, 1.5, 2, 3)
sens_results <- lapply(gammas, function(g) {
  general_BART_transportability(pate_result, gamma = g)
})

```

```{r sensitivity-curve-results}
sens_results <- readRDS(system.file("extdata", "sensitivity_curve_results.rds", package = "princeBART"))
gammas <- c(1.1, 1.25, 1.5, 2, 3)
# Extract bounds
lower_bounds <- sapply(sens_results, function(x) x$lower$estimate)
upper_bounds <- sapply(sens_results, function(x) x$upper$estimate)

# Plot sensitivity curve
plot(gammas, upper_bounds, type = "l", col = "red", 
     ylim = range(c(lower_bounds, upper_bounds)),
     xlab = "Gamma", ylab = "PATE Bounds", main = "Sensitivity Analysis")
lines(gammas, lower_bounds, col = "blue")
abline(h = 0, lty = 2)
abline(h = pate_result$pate, lty = 3)
legend("topright", c("Upper", "Lower", "Null", "Point est."), 
       col = c("red", "blue", "black", "black"), lty = c(1, 1, 2, 3))
```




## References

- Ding, P., & Lu, J. (2017). Principal stratification analysis using principal scores. *Journal of the Royal Statistical Society: Series B*.
- Chipman, H. A., George, E. I., & McCulloch, R. E. (2010). BART: Bayesian additive regression trees. *The Annals of Applied Statistics*.
- Crump, R. K., Hotz, V. J., Imbens, G. W., & Mitnik, O. A. (2009). Dealing with limited overlap in estimation of average treatment effects. *Biometrika*.
